PapeR Brain

### 2022-01-05 Markdown/PapeR Brain.Rmd
Version 3 of Markdown
Markdown v4.0

## The neuron

The graphs you'll see here are a digital iteration of a project I started in 2021, aptly called "Paper Brain." I set up basic neuron models on paper which consists of only two actions:

@@

Evidently, this is an oversimplification for neuronal activity (no clear exploration of how networks change over time, neuromodulation, or all that jazz). However, it can be a pretty nifty tool to learn how to explore the grey areas between what we call "neural networks" in a computational perspective, and a biological one.

@@

To begin, I started by exploring how we make decisions. In the context of code, we love to rely on certain models to give our programs functionality. In effect, the code we use, when we use `if`, `ifelse`, or `else` statements, gives our programs autonomous "decision-making" capacity (if only in any basic sense).

@@

Something something something diagram something something something insert paper brain image

@@

```{r}
library(networkD3)
```

@@

@@

(same color can be confusing, label excitatory / inhibitory / input / output neurons?)

@@

(not compared to microscopy yet)

(arrow-paths to show neural transmission)

>>>>

## The Neuron

The graphs you'll see here are a digital iteration of a project I started in mid-2021, called "Paper Brain." I set up basic neuron models on paper which consist of only two actions:

@@

Furthermore, with this model, all neuronal connections are assumed to be unidirectional — between a presynaptic neuron and a postsynaptic neuron, a pulse will only ever propagate from presynaptic to postsynaptic neuron.

Evidently, this is an oversimplification for neuronal activity (no clear exploration of how networks change over time, variations in neurotransmitters, or neuromodulation). However, it can be a pretty nifty tool to learn how to explore the grey areas between what we call "neural networks" in a computational perspective, and a biological one.

@@

To begin, I started by exploring how we make decisions. In the context of code, we can rely on certain structures to give our programs functionality. In effect, the code we use, when we use `if`, `ifelse`, or `else` statements, gives our programs autonomous "decision-making" capacity (if only in any basic sense).

@@

The paper representations of the structures that will be generated below can be found in my [PaperBrain repository on GitHub](https://github.com/osimon81/PaperBrain), under the *Images* folder. Feel free to excuse the scribbles :)

@@

```{r}
library(networkD3)
library(readr)
```

@@

# Pipelining

## Linear MicroNetwork Pipeline

The code below is an adaptation of the function `constructLinearMicroNetwork()` which will construct a MicroNetwork formed from multiple MicroNetworks. The function will essentially link a pair of input and output neurons for two adjacent MicroNetworks by building a connection between the two. To test out the function, run `constructLinearMicroNetwork()` in the console. The code to run this function can also be found under this line in the .Rmd file.

```{r constructLinearMicroNetwork() function, eval=FALSE, include=FALSE}
constructLinearMicroNetwork <- function() {
  print("Hello! Welcome to version 1 of the concerted MicroNetwork construction interface.")
  print("This project is essentially a pipeline which allows for fluid force-directed network construction.")
  print("To create MicroNetworks, use the following lines as syntax examples:")
  print("AND(n,n): create an AND MicroNetwork with any number of input neurons n and any output neurons n.")
  print("NOT(3, 1): create a NOT MicroNetwork with 3 input neurons and 1 output neuron")
  circuitList <- c()

  src <- c()
  target <- c()

  id <- 1
  editMode <- TRUE
  while (editMode == TRUE) {

    selectedNet <- readline(prompt = "Name the MicroNetwork you would like to add: ")
    if (grepl("AND", selectedNet) == TRUE) {
      circuitList <- c(circuitList, selectedNet)
      identifier <- as.character(id)

      in_out_count <- strsplit(selectedNet, ",")
      inputs <- as.numeric(parse_number(in_out_count[[1]][1]))
      outputs <- as.numeric(parse_number(in_out_count[[1]][2]))
      print("MicroNetwork Type: And") # Unique feedback for AND MicroNetworks
      print(paste("MicroNetwork Identifier:", identifier))
      print(paste("MicroNetwork", identifier, "Inputs:", as.character(inputs)))
      print(paste("MicroNetwork", identifier, "Outputs:", as.character(outputs)))

      # Assemble AND MicroNetwork neurons here (for the time being, this network only makes computational sense when input number is set to 3)

        src <- c(src,

                 paste0(identifier, "(1a)+"),
                 paste0(identifier, "(2a)+"),
                 paste0(identifier, "(3a)+"),
                 paste0(identifier, "(1a)+"),
                 paste0(identifier, "(2a)+"),
                 paste0(identifier, "(3a)+"),
                 paste0(identifier, "(4a)+"),
                 paste0(identifier, "(4a)+"),
                 paste0(identifier, "(6a)-"),
                 paste0(identifier, "(7a)-"))

        target <- c(target,
                    paste0(identifier, "(4a)+"),
                    paste0(identifier, "(4a)+"),
                    paste0(identifier, "(4a)+"),
                    paste0(identifier, "(5a)+"),
                    paste0(identifier, "(5a)+"),
                    paste0(identifier, "(5a)+"),
                    paste0(identifier, "(6a)-"),
                    paste0(identifier, "(7a)-"),
                    paste0(identifier, "(5a)+"),
                    paste0(identifier, "(5a)+"))

        if (as.numeric(identifier) > 1) {
          src <- c(src,
                   paste0(as.character((as.numeric(identifier)-1)), "(Out1)"))

          target <- c(target,
                      paste0(identifier, "(In1)"))
        }

        for (i in 1:inputs) {
          src <- c(src,
                    paste0(identifier, "(In", as.character(i), ")"))

          target <- c(target,
                      paste0(identifier, "(", i, "a)+"))
         }

        for (i in 1:outputs) {
          src <- c(src,
                   paste0(identifier, "(5a)+"))

          target <- c(target,
                      paste0(identifier, "(Out", as.character(i), ")"))
        }
      }
    else if (grepl("LOOP", selectedNet) == TRUE) {
      circuitList <- c(circuitList, selectedNet)
      identifier <- as.character(id)

      in_out_count <- strsplit(selectedNet, ",")
      inputs <- parse_number(in_out_count[[1]][1])
      outputs <- parse_number(in_out_count[[1]][2])
      print("MicroNetwork Type: Loop") # Unique feedback for LOOP MicroNetworks
      print(paste("MicroNetwork Identifier:", identifier))
      print(paste("MicroNetwork", identifier, "Inputs:", inputs))
      print(paste("MicroNetwork", identifier, "Outputs:", outputs))

      # Assemble LOOP MicroNetwork neurons here:

        src <- c(src,
                 paste0(identifier, "(1l)+"),
                 paste0(identifier, "(2l)+"),
                 paste0(identifier, "(3l)+"),
                 paste0(identifier, "(4l)+"))

        target <- c(target,
                    paste0(identifier, "(2l)+"),
                    paste0(identifier, "(3l)+"),
                    paste0(identifier, "(4l)+"),
                    paste0(identifier, "(1l)+"))

        if(as.numeric(identifier) > 1) {
          src <- c(src,
                   paste0(as.character((as.numeric(identifier)-1)), "(Out", as.character(i), ")"))

          target <- c(target,
                      paste0(identifier, "(In1)"))
        }

        for (i in 1:inputs) {
          src <- c(src,
                   paste0(identifier, "(In", as.character(i), ")"))

          target <- c(target,
                      paste0(identifier, "(1l)+"))
        }

        for (i in 1:outputs) {
          src <- c(src,
                   paste0(identifier, "(3l)+"))

          target <- c(target,
                      paste0(identifier, "(Out", as.character(i), ")"))
        }
      }
    else if (grepl("NOT", selectedNet) == TRUE) {
      circuitList <- c(circuitList, selectedNet)
      identifier <- as.character(id)

      in_out_count <- strsplit(selectedNet, ",")
      inputs <- parse_number(in_out_count[[1]][1])
      outputs <- parse_number(in_out_count[[1]][2])
      print("MicroNetwork Type: NOT") # Unique feedback for NOT MicroNetworks
      print(paste("MicroNetwork Identifier:", identifier))
      print(paste("MicroNetwork", identifier, "Inputs:", inputs))
      print(paste("MicroNetwork", identifier, "Outputs:", outputs))

      # Assemble NOT MicroNetwork neurons here:

        src <- c(src,
                 paste0(identifier, "(1n)-"))

        target <- c(target,
                    paste0(identifier, "(2n)+"))

      if (as.numeric(identifier) > 1) {
        src <- c(src,
                 paste0(as.character((as.numeric(identifier)-1)), "(Out", as.character(i), ")"))

        target <- c(target,
                    paste0(identifier, "(In1)"))
      }

      for (i in 1:inputs) {
        src <- c(src,
                 paste0(identifier, "(In", as.character(i), ")"))

        target <- c(target,
                    paste0(identifier, "(1n)-"))
      }

      for (i in 1:outputs) {
        src <- c(src,
                 paste0(identifier, "(2n)+"))

        target <- c(target,
                    paste0(identifier, "(Out", as.character(i), ")"))
      }
    }
    else {
      print("You have entered an invalid MicroNetwork type.")
      id <- id - 1
    }

    continue <- readline(prompt = "Would you like to add a new MicroNetwork? (Y/N) ")

    if (toupper(continue) == "Y") {
      editMode <- TRUE
      id <- id + 1
    }
    else if (toupper(continue) == "N") {
      editMode <- FALSE
    }
    else {
      editMode <- FALSE
    }
  }

  networkDF <- data.frame(src, target)
  simpleNetwork(networkDF, fontSize = 14,
                linkDistance = 100, zoom = TRUE)
}

constructLinearMicroNetwork()
```

The nomenclature system can appear complex, but the system exists to give every neuron a unique identification which reveals how they operate and where they are located in a much larger network. The following function, `interpretNeuron(id)` exists to automatically decipher these notation systems for ease of understanding the properties of each neuron.

For instance, the function can decipher the IDs "1(2n)-" and "52(Out3)" as the outputs show below:

```{r interpretNeuron() function, echo=FALSE}
interpretNeuron <- function(id) {
  if (grepl("Out", id) == TRUE) {
    neuronAttributes <- list(
      "Neuron Type" = "Output Neuron (sensory)",
      "Activity" = "Excitatory",
      "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                           1),
      "Subpathway Identifier" = sub(".*Out", "", substr(id, 1, nchar(id) -
                                                          1))
    )
  }
  else if (grepl("In", id) == TRUE) {
    neuronAttributes <- list(
      "Neuron Type" = "Input Neuron (sensory)",
      "Activity" = "Excitatory",
      "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                           1),
      "Subpathway Identifier" = sub(".*In", "", substr(id, 1, nchar(id) -
                                                         1))
    )
  }
  else if (grepl("a)", id) == TRUE) {
    neuronAttributes <-
      list("Neuron Type" = "Interneuron (AND MicroNetwork)")
    if (grepl("-|[\U{2013}]", id) == TRUE) {
      neuronAttributes <- c(
        neuronAttributes,
        "Activity" = "Inhibitory",
        "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                             1),
        "Subpathway Identifier" = as.character(parse_number(substr(
          id, 2, nchar(id)
        )))
      )
    }
    else if (grepl(as.character("+"), id) == TRUE) {
      neuronAttributes <- c(
        neuronAttributes,
        "Activity" = "Excitatory",
        "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                             1),
        "Subpathway Identifier" = as.character(parse_number(substr(id, 2, nchar(id))))
      )
    }
  }
  else if (grepl("l)", id) == TRUE) {
    neuronAttributes <-
      list("Neuron Type" = "Interneuron (LOOP MicroNetwork)")
    if (grepl("-|[\U{2013}]", id) == TRUE) {
      neuronAttributes <- c(
        neuronAttributes,
        "Activity" = "Inhibitory",
        "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                             1),
        "Subpathway Identifier" = as.character(parse_number(substr(
          id, 2, nchar(id)
        )))
      )
    }
    else if (grepl(as.character("+"), id) == TRUE) {
      neuronAttributes <- c(
        neuronAttributes,
        "Activity" = "Excitatory",
        "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                             1),
        "Subpathway Identifier" = as.character(parse_number(substr(id, 2, nchar(id))))
      )
    }
  }
  else if (grepl("n)", id) == TRUE) {
    neuronAttributes <-
      list("Neuron Type" = "Interneuron (NOT MicroNetwork)")
    if (grepl("-|[\U{2013}]", id) == TRUE) {
      neuronAttributes <- c(
        neuronAttributes,
        "Activity" = "Inhibitory",
        "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                             1),
        "Subpathway Identifier" = as.character(parse_number(substr(
          id, 2, nchar(id)
        )))
      )
    }
    else if (grepl(as.character("+"), id) == TRUE) {
      neuronAttributes <- c(
        neuronAttributes,
        "Activity" = "Excitatory",
        "MicroNetwork Identifier" = substr(id, 1, regexpr("(", id, fixed = TRUE) -
                                             1),
        "Subpathway Identifier" = as.character(parse_number(substr(id, 2, nchar(id))))
      )
    }
  }

  str(neuronAttributes)
}

interpretNeuron(id = "1(2n)-")
interpretNeuron(id = "52(Out3)")
```

This utility function can help to easily identify the properties of particular neurons based on a relatively compact identification system.

The lists displayed here determine the following properties:

- Neuron Type (Input / Output / Interneuron)
- Neuronal Activity (Excitatory / Inhibitory)
- MicroNetwork Identifier (Which net this neuron is a part of in the sequence of MicroNetworks)
- Subpathway Identifier (Where this neuron is in the particular MicroNetwork)

Saving all the properties of neurons created (as well as their linkages from the source-target data frame) for future processing may serve a useful feature for further analysis in modeling.

@@

Use of the `simpleNetwork()` function allows for flexible creation of force-directed representations of biological neural networks. However, some limitations in the function narrow the scope of how these models can be made easier to understand. While `interpretNeuron()` can be run on every neuron to determine the activity of each neuron, setting up a color scheme for labeling distinct colors for excitatory, inhibitory, input or output neurons.

Counteracting this limitation may be achieved by using the `forceNetwork()` function, which grants more customization options for force-directed networks, including colors, and is a future direction encouraged for this project.

Examining the network as simply a force-directed network in isolation, without reference to the list of source and associated target neurons, can make it unintuitive to establish the directionality of the signal (from input neurons to interneurons to output neurons).

Since connections between a source (presynaptic) neuron and a target (postsynaptic) neuron are assumed to be unidirectional, potentially changing the ambiguous lines represented in typical force-directed networks to unidirectional arrows may clarify the ambiguity.

@@

Recent advances in the field of [_in vivo_ calcium imaging](https://www.inscopix.com/miniature-microscope-technology) have allowed for the opportunity to inspect synapses at the microscale, and by extension, neural circuits. Inspection of circuits and firing patterns for an array of model organisms can inform how to update the structures of previously-established MicroNetworks to anatomically-correct representations.

Cross-matching the connectivity of these modules across various regions in an organism, accessed through minimicroscopes may allow updated pathways to be generalizable across an entire brain for that organism.
